---
title: "Peer-graded Assignment: Course Project 1 - Reproducible Research"
author: "Karnam Vasudeva Rao"
date: "4/21/2020"
output: html_document
---



```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)
library(ggplot2)
```


## 1. Read activity data
```{r, warning=FALSE, message=FALSE}
# Read activity data
activity_data <- read.csv("~/activity.csv", stringsAsFactors = F, header = T)

# Assign date class to date column
activity_data$date <- as.Date(activity_data$date)

```

## 1. Explore data
```{r, warning=FALSE, message=FALSE}
# Check dimensions
dim(activity_data)
summary(activity_data)
knitr::kable(activity_data[1:10,])
```


## 2. Histogram of the total number of steps taken each day
```{r, warning=FALSE, message=FALSE}

# Remove rows with missing data
activity_data_noNA <- na.omit(activity_data)

activity_data_noNA$steps <-  as.numeric(activity_data_noNA$steps)

# Group data by date and find total steps per day.
activity_sum_steps <- 
  activity_data_noNA %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = T))


# plot histogram
hist(activity_sum_steps$total_steps, main="Total number of steps taken each day", 
     xlab="Total steps per day", col = 'dark blue')

```


## 3. Mean and median number of steps taken each day
```{r, warning=FALSE, message=FALSE}
median(activity_sum_steps$total_steps)
mean(activity_sum_steps$total_steps)
```

## 4. Time series plot of the average number of steps taken

```{r, warning=FALSE, message=FALSE}

# Group data by interval and find average steps per day.
activity_mean_steps <- 
  activity_data_noNA %>%
  group_by(interval) %>%
  summarize(total_steps = mean(steps, na.rm = TRUE))

# Plot time series plot of the average number of steps taken
ggplot(aes(x=interval,y=total_steps),data=activity_mean_steps)+geom_line(color = "orange", size = 1.0) +
   ggtitle("The average number of steps by interval") 

```


## 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r, warning=FALSE, message=FALSE}

# Find maximum average 5 min interval
activity_mean_steps[activity_mean_steps$total_steps == max(activity_mean_steps$total_steps),1]

```


## 6. Code to describe and show a strategy for imputing missing data
```{r, warning=FALSE, message=FALSE}

# Check missing values in all the three columns (as there are only three columns)
sum(is.na(activity_data$steps))
sum(is.na(activity_data$date))
sum(is.na(activity_data$interval))

# Data is missing in 2304 rows in steps column

# Fill missing rows with column mean value
activity_data$steps[is.na(activity_data$steps)] <- mean(activity_data$steps,na.rm=TRUE)

```


## 7. Histogram of the total number of steps taken each day after missing values are imputed
```{r, warning=FALSE, message=FALSE}

# Find total number of steps taken per day after imputing
activity_sum_steps_imputed <- 
  activity_data %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps, na.rm = T))

# Plot histogram
hist(activity_sum_steps_imputed$total_steps, main="Total number of steps taken each day (post imputing)", 
     xlab="Total steps per day", col = 'dark green')

```

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r, warning=FALSE, message= FALSE}

# Map day name for date
activity_data$days=tolower(weekdays(activity_data$date))

# head(activity_data)
# unique(activity_data$days)

# Assign days to weekend or weekday
activity_data$category <- ifelse(grepl("saturday|sunday", activity_data$days),"weekend","weekday")

# Find average number of steps on weekend or weekday in the intervals
activity_mean_steps_days <- 
  activity_data %>%
  group_by(interval, category) %>%
  summarize(mean_steps = mean(steps, na.rm = T))


# Plot average steps and interval as per day category

ggplot(aes(x=interval,y=mean_steps),data=activity_mean_steps_days)+geom_line(color = "dark blue", size = 1.0)+facet_wrap(~activity_mean_steps_days$category)

```

